# MPB maps data processing

**Alex M. Chubaty**

Canadian Forest Service
Pacific Forestry Centre
506 Burnside Road W
Victoria, BC V8Z 1M5

Université Laval
Faculté de foresterie, de géographie et de géomatique
Département des sciences du bois et de la forêt

*email:* [Alexander.Chubaty@nrcan.gc.ca][1]
*phone:* +1.250.298.2347

[1]: <mailto:Alexander.Chubaty@nrcan.gc.ca>

------------------------------------------------------------------

## Overview of MPB map data

Descriptions of:

- data sources
- data collected
- etc.

```{r setup.workspace, echo=FALSE, message=FALSE}
### SETUP R WORKING ENVIROMENT
source("workspace.maps.R")

num.cpus = 4            # maximum cpus to use

read.in.raw.maps = FALSE
reproj.raw.maps = FALSE
rasterize.maps = FALSE

res.maps = 1000
ext.maps = extent(x=-1027658, xmax=320751.9, ymin=5108872, ymax=6163350)
# map extent obtained using `locator(2)` on 'west' map
```

## Importing map data

### MPB `SpatialPoints` data

```{r import.maps-points, echo=FALSE}
if (read.in.raw.maps) {
  # Read in raw maps, and save them as individual files
  # - this may take a while...grab a coffee or something
  source(file.path(work.dir, "import.raw.maps-points.R"))
}
```

### MPB `SpatialPolygons` data

```{r import.maps-polygons, echo=FALSE}
if (read.in.raw.maps) {
  # Read in raw maps, and save them as individual files
  # - this may take a while...grab a coffee or something
  source(file.path(work.dir, "import.raw.maps-polygons.R"))
}
```

### Canadian boreal forest maps

```{r import.maps-boreal, echo=FALSE}
if (read.in.raw.maps) {
  # Read in raw maps, and save them as individual files
  # - this may take a while...grab a coffee or something
  source(file.path(work.dir, "import.raw.maps-boreal.R"))
}
```

Boreal forest map data uses the `crs.boreal` projection below. We reproject the Canadian administrative boundaries maps to use this projection:

```{r crs.boreal}
if(!exists("crs.boreal")) {
  loadObjects("boreal", rdata.path)
  crs.boreal = CRS(proj4string(boreal))
  rm(boreal)
}
```

**The Canadian boreal forest:**

```{r plot.canada.boreal fig.height=4 fig.width=7}
loadObjects(c("boreal.can", "canada1.boreal"), rdata.path)
darkgreen <- col2rgb("darkgreen")/255
col.bor <- rgb(darkgreen[1], darkgreen[2], darkgreen[3], alpha=0.5)
rm(darkgreen)

plot(canada1.boreal)
plot(boreal.can, col=col.bor, add=TRUE)
rm(boreal.can)
```

**The western Canadian boreal forest:**

```{r plot.western.boreal fig.height=5, fig.width=6}
loadObjects(c("boreal.west", "west.boreal"), rdata.path)
darkgreen <- col2rgb("darkgreen")/255
col.bor <- rgb(darkgreen[1], darkgreen[2], darkgreen[3], alpha=0.5)
rm(darkgreen)

plot(west.boreal)
plot(boreal.west, col=col.bor, add=TRUE))
rm(boreal.west)
```

## Reproject MPB maps to `crs.boreal`

### MPB `SpatialPoints` data

```{r reproject.maps-points echo=FALSE}
if (reproj.raw.maps) {
  loadObjects(c("ab.pnts", "bc.pnts"), rdata.path)
  source(file.path(work.dir, "reproject.maps-points.R"))
  source(file.path(work.dir, "merge.maps-points.R"))
  rm(ab.pnts, bc.pnts)
}
```

### MPB `SpatialPolygons` data

```{r reproject.maps-polygons echo=FALSE}
if (reproj.raw.maps) {
  loadObjects(c("ab.poly", "bc.poly"), rdata.path)
  source(file.path(work.dir, "reproject.maps-polygons.R"))
  # merging poly data can't be done at this point -- will merge rasters
  rm(ab.poly, bc.poly)
}
```

## Rasterize MPB maps

### Canadian boreal forest maps

```{r rasterize.maps-boreal echo=FALSE}
if (rasterize.maps) {
  loadObjects(c("bcab.pnts.boreal", "west.boreal.raster"), rdata.path)
  source(file.path(work.dir, "rasterize.maps-boreal.R"))
  rm(bcab.pnts.boreal, west.boreal)
}
```

### MPB `SpatialPoints` data

```{r rasterize.maps-points echo=FALSE}
if (rasterize.maps) {
  loadObjects(c("ab.pnts.boreal", "bc.pnts.boreal", "west.boreal.raster"), rdata.path)
  source(file.path(work.dir, "rasterize.maps-points.R"))
  rm(ab.pnts.boreal, bc.pnts.boreal, west.boreal.raster)
}
```

### MPB `SpatialPolygons` data

```{r rasterize.maps-polygons echo=FALSE}
if (rasterize.maps) {
  loadObjects(c("ab.poly.boreal", "bc.poly.boreal", "west.boreal.raster"), rdata.path)
  source(file.path(work.dir, "rasterize.maps-polygons.R"))
  source(file.path(work.dir, "merge.maps-polygons.R"))
  rm(ab.poly.boreal, bc.poly.boreal, west.boreal.raster)
}
```

