---
title: "MPB dynamics and spread"
author:
  - "Alex M. Chubaty"
  - "Barry J. Cooke"
  - "Eliot J. B. McIntire"
date: "February 05, 2018"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
bibliography:
  - bibliography.bib
  - ../mpb-maps/climate-projections.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE, eval = TRUE, message = FALSE,
                      warning = FALSE)

## make sure CRAN mirror is set
if (is.null(getOption("repos"))) {
  options(repos = c(CRAN = "https://cran.rstudio.com/"))
}

## install/load packages
pkgs <- c("data.table", "devtools", "ggplot2", "igraph", "magrittr", "plyr",
          "scales", "sp", "rgdal", "raster",
          "achubaty/amc@development",
          "PredictiveEcology/LandR@development",
          #"PredictiveEcology/map@development",
          #"PredictiveEcology/pemisc@development",
          "PredictiveEcology/reproducible@development",
          "PredictiveEcology/SpaDES.core@development",
          "PredictiveEcology/SpaDES.tools@development"
)

reproducible::Require(pkgs)
```

# Introduction

The goal of this model is to simulate the spread of mountain pine beetle outbreaks as a function of the growth and dispersal of beetle-killed trees, called "red tops".

## Study area

Our study area spans a region of north-central Alberta (AB) and into Saskatchewan (SK), Canada (114.9704°W – 108.4075°W; 54.42415°N – 56.35593°N).
This area was selected because it coincides closely with the Cold Lake Air Weapons Range, which represents the east-most extension of MPB in boreal jack pine.
The west end extends to Lesser Slave Lake, so there are light populations persistently established since 2010, which yields enough data to test model growth and spread predictions.

**NOTE:** major GIS operations such as reprojecting and cropping are cached, to speed up subsequent running of the model.

```{r define-study-area}
## Canadian administrative boundaries
admPath <- checkPath("~/data/CAN_adm", create = TRUE)
CAN_adm1 <- getData("GADM", country = "CAN", level = 1, path = admPath)
CAN_adm3 <- getData("GADM", country = "CAN", level = 3, path = admPath)

ab <- CAN_adm1[CAN_adm1$NAME_1 == "Alberta", ]
ab3 <- CAN_adm3[CAN_adm3$NAME_1 == "Alberta", ]
ab_cities <- ab3[ab3$ENGTYPE_3 == "City", ]
bc <- CAN_adm1[CAN_adm1$NAME_1 == "British Columbia", ]
west <- CAN_adm1[(CAN_adm1$NAME_1 == "British Columbia" |
                  CAN_adm1$NAME_1 == "Alberta" |
                  CAN_adm1$NAME_1 == "Saskatchewan"), ]

## load studyArea
prj <- paste("+proj=aea +lat_1=47.5 +lat_2=54.5 +lat_0=0 +lon_0=-113",
             "+x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0")
f <- normalizePath("~/GitHub/MPB/studyArea/studyArea.kml", mustWork = TRUE)
studyArea <- readOGR(f, "studyArea.kml") %>%
  spTransform(., prj)
west2 <- spTransform(west, proj4string(studyArea))

studyArea@data$id <- rownames(studyArea@data)
studyArea_pnts <- fortify(studyArea, region = "id")
studyArea_df <- join(studyArea_pnts, studyArea@data, by = "id")

west2@data$id <- rownames(west2@data)
west2_pnts <- fortify(west2, region = "id")
west2_df <- join(west2_pnts, west2@data, by = "id")

ggplot(west2_df) +
  aes(long, lat, group = group) +
  geom_polygon() +
  geom_path(color = "white") +
  coord_equal() +
  geom_polygon(data = studyArea_df, color = "white", fill = hue_pal()(16)[11]) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.text.y = element_blank())
```

## Module descriptions

```{r child = 'modules/canWind/canWind.Rmd'}
```

```{r child = 'modules/mpbClimateData/mpbClimateData.Rmd'}
```

```{r child = 'modules/mpbMassAttacksData/mpbMassAttacksData.Rmd'}
```

```{r child = 'modules/mpbPine/mpbPine.Rmd'}
```

```{r child = 'modules/mpbRedTopGrowth/mpbRedTopGrowth.Rmd'}
```

```{r child = 'modules/mpbRedTopSpread/mpbRedTopSpread.Rmd'}
```

```{r child = 'modules/mpbManagement/mpbManagement.Rmd'}
```

# Initialize and run the simulation

## Simulation setup

```{r sim-setup}
projDir <- checkPath("~/GitHub/MPB")
setPaths(
  cachePath = file.path(projDir, "cache"),
  inputPath = file.path(projDir, "inputs"),
  modulePath = file.path(projDir, "modules"),
  outputPath = file.path(projDir, "outputs")
)

if (interactive()) dev.useRSGD(FALSE)

## set to TRUE on laptop
lowMemory <- ifelse(Sys.info()["nodename"] %in% c("landweb"), FALSE, TRUE)
maxMemory <- 5e+12
scratchDir <- file.path("/tmp/scratch/MPB")

rasterOptions(default = TRUE)
options(rasterMaxMemory = maxMemory, rasterTmpDir = scratchDir)

times <- list(start = 2011, end = 2020)
params <- list(
  mpbClimateData = list(
    suitabilityIndex = "G",    ## Can be "G", "S", "L", "R"
    .maxMemory = maxMemory,
    .tempdir = scratchDir
  ),
  mpbMassAttacksData = list(
    .maxMemory = maxMemory,
    .tempdir = scratchDir
  ),
  mpbPine = list(
    lowMemory = lowMemory,
    .maxMemory = maxMemory,
    .tempdir = scratchDir
  ),
  mpbRedTopGrowth = list(
    dataset = "Boone_2011"
  )
)
#objects <- list(studyArea = studyArea)

## use EITHER mpbRandomLandscapes OR mpbPine/mpbClimateData (not all three)
modules <- list(
  #"mpbRandomLandscapes",
  "mpbClimateData","mpbPine",
  "mpbMassAttacksData",
  "mpbRedTopGrowth",
  "mpbRedTopSpread",
  "mpbManagement"
)

MPB <- simInit(times = times, params = params, modules = modules, #objects = objects,
               loadOrder = unlist(modules))
```

## Run the model

```{r simulation, fig.keep="last", eval=FALSE}
## the first time you run this, esp. w/ large study area, it will be slow because
## of the GIS operations taking place. However, because these slow steps are cached,
## subsequent runs will be much faster (plotting will became slowest step).
MPB2 <- spades(Copy(MPB), debug = TRUE)
```

## Simulation diagrams

### Module diagram

```{r module-diagram, fig.keep="last", eval=FALSE}
clearPlot()
vcol <- sapply(names(V(depsGraph(MPB))), function(v) {
  if (v == "_INPUT_") {
    "orange"
  } else if (v %in% c("mpbRedTopGrowth", "mpbRedTopSpread", "mpbManagement")) {
    "pink"
  } else {
    "lightblue"
  }
})
moduleDiagram(MPB, vertex.color = vcol, vertex.size = 30)
```

### Object diagram

```{r object-diagram, fig.keep="last", eval=FALSE}
clearPlot()
objectDiagram(MPB)
```

### Event diagram

```{r event-diagram, fig.keep="last", eval=FALSE}
clearPlot()
eventDiagram(MPB2)
```

# References

<!-- automatically generated from the bibbligrapty files in the yaml header above -->
