---
title: "MPB dynamics and spread"
author:
  - "Alex M. Chubaty"
  - "Barry J. Cooke"
  - "Eliot J. B. McIntire"
date: "February 05, 2018"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
bibliography:
  - bibliography.bib
  - mpb-maps/climate-projections.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE, eval = TRUE, message = FALSE,
                      warning = FALSE)

## make sure CRAN mirror is set
if (is.null(getOption("repos"))) {
  options(repos = c(CRAN = "https://cran.rstudio.com/"))
}

## install/load packages
#install_github("PredictiveEcology/reproducible@development")
#install_github("PredictiveEcology/SpaDES.core@development")
#install_github("achubaty@amc@development")

library(magrittr)
library(raster)
library(reproducible)
library(SpaDES.core)
library(amc)
library(ggplot2)
library(DEoptim)
```

# Introduction

The goal of this model is to simulate the spread of mountain pine beetle outbreaks as a function of the growth and dispersal of beetle-killed trees, called "red tops".

## Study area

Our study area spans a region of north-central Alberta (AB) and into Saskatchewan (SK), Canada (114.9704°W – 108.4075°W; 54.42415°N – 56.35593°N).
This area was selected because it coincides closely with the Cold Lake Air Weapons Range, which represents the east-most extension of MPB in boreal jack pine.
The west end extends to Lesser Slave Lake, so there are light populations persistently established since 2010, which yields enough data to test model growth and spread predictions.

**NOTE:** major GIS operations such as reprojecting and cropping are cached, to speed up subsequent running of the model.

```{r define-study-area}
## Canadian administrative boundaries
admPath <- checkPath("~/data/CAN_adm", create = TRUE)
CAN_adm1 <- getData("GADM", country = "CAN", level = 1, path = admPath)
CAN_adm3 <- getData("GADM", country = "CAN", level = 3, path = admPath)

ab <- CAN_adm1[CAN_adm1$NAME_1 == "Alberta", ]
ab3 <- CAN_adm3[CAN_adm3$NAME_1 == "Alberta", ]
ab_cities <- ab3[ab3$ENGTYPE_3 == "City", ]
bc <- CAN_adm1[CAN_adm1$NAME_1 == "British Columbia", ]
west <- CAN_adm1[(#CAN_adm1$NAME_1 == "British Columbia" |
                  CAN_adm1$NAME_1 == "Alberta" |
                  CAN_adm1$NAME_1 == "Saskatchewan"), ]

## load studyArea for plotting
prj <- paste("+proj=aea +lat_1=47.5 +lat_2=54.5 +lat_0=0 +lon_0=-113",
             "+x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0") ## NOTE: aea not lcc
f <- normalizePath("~/GitHub/MPB/studyArea/studyArea.kml", mustWork = TRUE)
studyArea <- readOGR(f, "studyArea.kml") %>% spTransform(., prj)
west2 <- spTransform(west, proj4string(studyArea))

studyArea@data$id <- rownames(studyArea@data)
studyArea_pnts <- fortify(studyArea, region = "id")
studyArea_df <- join(studyArea_pnts, studyArea@data, by = "id")

west2@data$id <- rownames(west2@data)
west2_pnts <- fortify(west2, region = "id")
west2_df <- join(west2_pnts, west2@data, by = "id")

ggplot(west2_df) +
  aes(long, lat, group = group) +
  geom_polygon() +
  geom_path(color = "white") +
  coord_equal() +
  geom_polygon(data = studyArea_df, color = "white", fill = hue_pal()(16)[11]) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.text.y = element_blank())
```

## Module descriptions

```{r child = 'modules/canWind/canWind.Rmd'}
```

```{r child = 'modules/mpbClimateData/mpbClimateData.Rmd'}
```

```{r child = 'modules/mpbMassAttacksData/mpbMassAttacksData.Rmd'}
```

```{r child = 'modules/mpbPine/mpbPine.Rmd'}
```

```{r child = 'modules/mpbRedTopGrowth/mpbRedTopGrowth.Rmd'}
```

```{r child = 'modules/mpbRedTopSpread/mpbRedTopSpread.Rmd'}
```

```{r child = 'modules/mpbManagement/mpbManagement.Rmd'}
```

# Initialize and run the simulation

## Simulation setup

```{r sim-setup}
projDir <- checkPath("~/GitHub/MPB")
scratchDir <- checkPath(file.path("/mnt/scratch/MPB"), create = TRUE)
setPaths(
  cachePath = file.path(projDir, "cache"),
  inputPath = file.path(projDir, "inputs"),
  modulePath = file.path(projDir, "modules"),
  outputPath = file.path(projDir, "outputs"),
  rasterPath = file.path(scratchDir, "rasters")
)

if (interactive()) dev.useRSGD(FALSE)

RenvironFile <- file.path(projDir, ".Renviron")
if (file.exists(RenvironFile)) readRenviron(RenvironFile)

cacheDBconn <- DBI::dbConnect(drv = RPostgres::Postgres(),
                              host = Sys.getenv("PGHOST"),
                              port = Sys.getenv("PGPORT"),
                              dbname = Sys.getenv("PGDATABASE"),
                              user = Sys.getenv("PGUSER"),
                              password = Sys.getenv("PGPASSWORD")) 
lowMemory <- ifelse(Sys.info()["nodename"] %in% c("forcast01"), FALSE, TRUE)
maxMemory <- 5e+12

raster::rasterOptions(default = TRUE)
options(
  "rasterMaxMemory" = maxMemory,
  "rasterTmpDir" = Paths$rasterPath,
  "reproducible.cacheSaveFormat" = "rds", ## can be "qs" or "rds"
  "reproducible.conn" = cacheDBconn
)

rasterOptions(default = TRUE)
options(rasterMaxMemory = maxMemory, rasterTmpDir = Paths$rasterPath)

timesFit <- list(start = 2010, end = 2011) ## 2010-2016
paramsFit <- list(
  mpbClimateData = list(
    suitabilityIndex = "R",    ## Can be "G", "S", "L", "R"
    .maxMemory = maxMemory,
    .tempdir = scratchDir
  ),
  mpbMassAttacksData = list(
    .maxMemory = maxMemory,
    .tempdir = scratchDir
  ),
  mpbPine = list(
    lowMemory = lowMemory,
    .maxMemory = maxMemory,
    .tempdir = scratchDir
  ),
  mpbRedTopGrowth = list(
    dataset = "Boone2011"
  ),
  mpbRedTopSpread = list(
    advectionDir = 90,
    advectionMag = 1000,
    bgSettlingProp = 0.1,
    meanDist = 1000
  )
)

#objects <- list(studyArea = studyArea) ## studyArea loaded by modules

## use EITHER mpbRandomLandscapes OR mpbPine/mpbClimateData (not all three)
modules <- list(
  #"mpbRandomLandscapes",
  "mpbClimateData", "mpbPine",
  "mpbMassAttacksData",
  "mpbRedTopGrowth",
  "mpbRedTopSpread",
  "mpbManagement"
)
```

## Run the model: fitting spread parameters

```{r fit-with-POM, fig.keep="last", eval=FALSE}
MPBfit <- Cache(simInit,
                times = timesFit,
                params = paramsFit,
                modules = modules,
                #objects = objects,
                loadOrder = unlist(modules)#,
                #useCache = "overwrite"
)

objectiveFunction <- function(params) {
  data.table::setDTthreads(1L)
  
  RenvironFile <- file.path(projDir, ".Renviron")
  if (file.exists(RenvironFile)) readRenviron(RenvironFile)
  
  cacheDBconn <- DBI::dbConnect(drv = RPostgres::Postgres(),
                                host = Sys.getenv("PGHOST"),
                                port = Sys.getenv("PGPORT"),
                                dbname = Sys.getenv("PGDATABASE"),
                                user = Sys.getenv("PGUSER"),
                                password = Sys.getenv("PGPASSWORD")) 

  SpaDES.core::setPaths(
    cachePath = file.path(projDir, "cache"),
    inputPath = file.path(projDir, "inputs"),
    modulePath = file.path(projDir, "modules"),
    outputPath = file.path(projDir, "outputs"),
    rasterPath = file.path(scratchDir, "rasters")
  )
  lowMemory <- TRUE
  maxMemory <- 5e+9
  
  raster::rasterOptions(default = TRUE)
  options(
    "rasterMaxMemory" = maxMemory,
    "rasterTmpDir" = Paths$rasterPath,
    "reproducible.cacheSaveFormat" = "rds", ## can be "qs" or "rds"
    "reproducible.conn" = cacheDBconn
  )

  Sys.sleep(runif(1, 1, 100)) ## random delay to mitigate database locked errors
  
  sim <- reproducible::retry(quote({
    simInit(times = timesFit, params = paramsFit, modules = modules,
            #objects = objects,
            #useCache = "overwrite",
            loadOrder = unlist(modules)
    )
  }))

  params(sim)[["mpbRedTopSpread"]][["advectionDir"]] <- params[1]
  params(sim)[["mpbRedTopSpread"]][["advectionMag"]] <- params[2]
  params(sim)[["mpbRedTopSpread"]][["bgSettlingProp"]] <- params[3]
  params(sim)[["mpbRedTopSpread"]][["meanDist"]] <- params[4]
  
  simOut <- spades(sim, .plotInitialTime = NA, debug = FALSE)
  
  ## TODO: below currently does total attack area. more refined criteria needed
  ## e.g., area and shape metrics from `landscapemetrics` package?
  
  ## simulated attack area
  atks <- simOut$massAttacksDT
  nPix <- atks[ATKTREES > 0, .N]
  atkAreaSim <- nPix * prod(res(simOut$rasterToMatch)) / (100^2) ## area in ha
  
  ## attacked area from data
  atksRas <- simOut$massAttacksMap[[paste0("X", timesFit$end)]]
  atks <- data.table(ID = 1L:ncell(atksRas), ATKTREES = atksRas[])
  nPix <- atks[ATKTREES > 0, .N] ## total number of pixels
  atkAreaData <- nPix * prod(res(simOut$rasterToMatch)) / (100^2) ## area in ha
  
  ## sum negative log likelihood for attacked pixels
  
  
  ## TODO: something other than simple sum of squares?
  metric <- (atkAreaData - atkAreaSim)^2 #+ (SNLL / 10^3)
  return(metric)
}

params4POM <- data.frame(
  name = c("advectionDir", "advectionMag", "bgSettlingProp", "meanDist"),
  lower = c(  0.000,   10, 0.01,   100),
  upper = c(359.999, 1000, 0.20, 10000), ## TODO: refine these upper and lower limits
  stringsAsFactors = FALSE
)

packages4POM <- lapply(unlist(modules), function(m) reqdPkgs(MPBfit, m)) %>%
  unlist() %>%
  unique() %>% 
  grep("@", ., invert = TRUE, value = TRUE) %>% 
  c("amc", "LandR", "pemisc", "SpaDES.core")

N <- 10 * nrow(params4POM) ## need 10 populations per parameter
cl <- parallel::makeCluster(min(N, 20),     ## up to 20 GB each for 2010-2011, so max 24 threads
                            type = "SOCK")  ## forking doesn't work with data.table

parallel::clusterExport(cl, varlist = c("modules", "paramsFit", "packages4POM", "projDir", "scratchDir", "timesFit"))
parallel::clusterEvalQ(
    cl, {
      for (i in packages4POM)
        library(i, character.only = TRUE)
    }
)
outPOM <- DEoptim(fn = objectiveFunction,
                  control = DEoptim::DEoptim.control(
                    cluster = cl, ## see ArdiaD/DEoptim#3
                    #foreachArgs = c(.packages = packages4POM), ## only when parallelType = 2
                    initialpop = NULL,
                    itermax = 50,
                    #packages = as.list(packages4POM),
                    parallelType = 1, ## 0 = single thread; 1 = parallel; 2 = foreach
                    #parVar = list("modules", "paramsFit", "timesFit", "Paths"),
                    VTR = 0
                  ),
                  lower = params4POM$lower,
                  upper = params4POM$upper
)
parallel::stopCluster(cl)

qs::qsave(outPOM, file.path(Paths$outputPath, "outPOM.qs"))
#saveSimList(MPBfit, file.path(Paths$outputPath, "test.qs"))
#MPBfit <- loadSimList(file.path(Paths$outputPath, "test.qs"))
```

## Run the model: prediction

```{r simulation, fig.keep="last", eval=FALSE}
timesPredict <- list(start = 2017, end = 2020)

paramsPredict <- paramsFit

bestFitVals <- outPOM[["optim"]][["bestmem"]]

paramsPredict[["mpbRedTopSpread"]][["advectionDir"]] <- bestFitVals[[1]]
paramsPredict[["mpbRedTopSpread"]][["advectionMag"]] <- bestFitVals[[2]]
paramsPredict[["mpbRedTopSpread"]][["bgSettlingProp"]] <- bestFitVals[[3]]
paramsPredict[["mpbRedTopSpread"]][["meanDist"]] <- bestFitVals[[4]]

MPBpred <- simInitAndSpades(times = timesPredict, params = paramsPredict, modules = modules,
                            #objects = objects,
                            loadOrder = unlist(modules))
#saveSimList(MPBpred, file.path(Paths$outputPath, "MPBpred.qs"))
#MPBpred <- loadSimList(file.path(Paths$outputPath, "MPBpred.qs"))
```

## Simulation diagrams

### Module diagram

```{r module-diagram, fig.keep="last", eval=FALSE}
clearPlot()
vcol <- sapply(names(V(depsGraph(MPB))), function(v) {
  if (v == "_INPUT_") {
    "orange"
  } else if (v %in% c("mpbRedTopGrowth", "mpbRedTopSpread", "mpbManagement")) {
    "pink"
  } else {
    "lightblue"
  }
})
moduleDiagram(MPB, vertex.color = vcol, vertex.size = 30)
```

### Object diagram

```{r object-diagram, fig.keep="last", eval=FALSE}
clearPlot()
objectDiagram(MPB)
```

### Event diagram

```{r event-diagram, fig.keep="last", eval=FALSE}
clearPlot()
eventDiagram(MPB2)
```

# References

<!-- automatically generated from the bibbligrapty files in the yaml header above -->
