---
title: "MPB dynamics and spread"
author:
  - "Alex M. Chubaty"
  - "Barry J. Cooke"
  - "Eliot J. B. McIntire"
date: "January 15, 2018"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
bibliography:
  - bibliography.bib
  - ../mpb-maps/climate-projections.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE, eval = TRUE, message = FALSE,
                      warning = FALSE)
```

# Introduction

The goal of this model is to simulate the spread of mountain pine beetle outbreaks as a function of the growth and dispersal of beetle-killed trees, called "red tops".

## Study area

Lorem ipsum...

## Module descriptions

```{r child = 'modules/canWind/canWind.Rmd'}
```

```{r child = 'modules/mpbClimateData/mpbClimateData.Rmd'}
```

```{r child = 'modules/mpbMassAttacksData/mpbMassAttacksData.Rmd'}
```

```{r child = 'modules/mpbPine/mpbPine.Rmd'}
```

```{r child = 'modules/mpbRedTopGrowth/mpbRedTopGrowth.Rmd'}
```

```{r child = 'modules/mpbRedTopSpread/mpbRedTopSpread.Rmd'}
```

```{r child = 'modules/mpbManagement/mpbManagement.Rmd'}
```

# Installation and setup

Install the most recent version of the `SpaDES` package from the development branch (packages version `2.0.0` or higher is required.)

```{r install-packages}
if (is.null(getOption("repos"))) {
  options(repos = c(CRAN = "https://cran.rstudio.com/"))
}

if (!require(data.table)) {
  install.packages("data.table")
}
if (!require(devtools)) {
  install.packages("devtools")
}
if (!require(igraph)) {
  install.packages("igraph")
}
if (!require(sp)) {
  install.packages("sp")
}
if (!require(rgdal)) {
  install.packages("rgdal")
}
if (!require(raster)) {
  install.packages("raster")
}
if (!require(amc) || packageVersion("amc") < "0.1.3") {
  devtools::install_github("achubaty/amc@development", dependencies = TRUE)
}
if (!require(SpaDES) || packageVersion("SpaDES") < "2.0.0") {
  devtools::install_github("PredictiveEcology/SpaDES@development", dependencies = TRUE)
}
```

Load required packages and set some default options for data loading etc.

```{r additional-setup}
setPaths(
  cachePath = "~/SpaDES/cache",
  modulePath = "~/GitHub/MPB/SpaDES/modules"
)
if (interactive()) dev.useRSGD(FALSE)
```

# Initialize and run the simulation

## Define study area

The study area straddles the Alberta (AB) and Saskatchewan (SK) border between Lesser Slave Lake to the west and Cold Lake to the east, per discussions with Barry.
This area includes the Martin Hills in AB.

**NOTE:** major GIS operations such as reprojecting and cropping are cached, to speed up subsequent running of the model.

```{r define-study-area, eval=FALSE}
## load CRS for the boreal raster
load(file.path(getOption("spades.modulePath"), "mpbMassAttacksData", "data",
               "west.boreal.RData")) ## loads studyArea object (spdf)
crs.boreal <- CRS(proj4string(studyArea))
rm(studyArea)

## load studyArea
f <- normalizePath("studyArea/studyArea.kml")
#ogrInfo(f)
#ogrListLayers(f)
studyArea <- readOGR(f, "studyArea.kml")

if (!file.exists(extension(f, ".shp"))) {
  writeOGR(studyArea, dsn = extension(f, ".shp"), layer = "studyArea.kml", driver = "ESRI Shapefile")
}

times <- list(start = 2011, end = 2020)
params <- list(
  mpbClimateData = list(suitabilityIndex = "G"),    ## Can be "G", "S", "L", "R"
  mpbPine = list(lowMemory = TRUE), ## set to TRUE on laptop
  mpbRedTopGrowth = list(dataset = "Berryman1979_forced")
)
objects <- list(studyArea = demoArea)

## use EITHER mpbRandomLandscapes OR mpbPine/mpbClimateData (not all three)
modules <- list(
  #"mpbRandomLandscapes",
  "mpbPine", "mpbClimateData",
  "mpbMassAttacksData",
  "mpbRedTopGrowth",
  "mpbRedTopSpread",
  "mpbManagement"
)
```

## Run the model

```{r simulation, fig.keep="last", eval=FALSE}
MPB <- simInit(times = times, params = params, modules = modules, objects = objects)

## the first time you run this, esp. w/ large study area, it will be slow because
## of the GIS operations taking place. However, because these slow steps are cached,
## subsequent runs will be much faster (plotting will became slowest step).
MPB2 <- spades(Copy(MPB), debug = TRUE)
```

## Simulation diagrams

### Module diagram

```{r module-diagram, fig.keep="last", eval=FALSE}
clearPlot()
vcol <- sapply(names(V(depsGraph(MPB))), function(v) {
  if (v == "_INPUT_") {
    "orange"
  } else if (v %in% c("mpbRedTopGrowth", "mpbRedTopSpread", "mpbManagement")) {
    "pink"
  } else {
    "lightblue"
  }
})
moduleDiagram(MPB, vertex.color = vcol, vertex.size = 30)
```

### Object diagram

```{r object-diagram, fig.keep="last", eval=FALSE}
clearPlot()
objectDiagram(MPB)
```

### Event diagram

```{r event-diagram, fig.keep="last", eval=FALSE}
clearPlot()
eventDiagram(MPB2)
```

# References

<!-- automatically generated from the bibbligrapty files in the yaml header above -->
