---
title: "Pine maps"
author: "Alex Chubaty"
date: "May 25, 2021"
output: pdf_document
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = FALSE, eval = TRUE)
```

```{r packages-setup}
library("Require")
Require(c("httr", "magrittr", "raster", "rasterVis", "RColorBrewer", "RCurl", "reproducible"))
Require(c("achubaty/amc@development",
          "PredictiveEcology/LandR@development",
          "PredictiveEcology/mpbutils"))

cacheDir <- checkPath("cache", create = TRUE)
inputsDir <- checkPath("inputs", create = TRUE)
scratchDir <- checkPath(file.path(tempdir(), "scratch"), create = TRUE)

rasterOptions(default = TRUE)
options(
  rasterTmpDir = scratchDir,
  reproducible.cachePath = cacheDir
)

myThemeGreen <- rasterTheme(region = brewer.pal("Greens", n = 9)[3:9])
myThemeBlue <- rasterTheme(region = brewer.pal("Blues", n = 9))

targetCRS <- paste("+proj=lcc +lat_1=49 +lat_2=77 +lat_0=0 +lon_0=-95",
                   "+x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
studyArea <- mpbStudyArea(ecoregions = c(112, 120, 122, 124, 126), targetCRS, cacheDir, inputsDir)
rasterToMatch <- Cache(
  LandR::prepInputsLCC,
  year = 2005, ## TODO: use 2010
  destinationPath = inputsDir,
  studyArea = sf::as_Spatial(studyArea)
)
```

# Comparison of pine maps from different sources

TODO: need summary/overview

## Yemshanov pine map

From @Yemshanov:2012ea (a copy of the paper is included with the dataset downloaded by the code below).

```{r yemshanov-plot}
url_yemshanov <- "https://drive.google.com/file/d/11g02bDnEt6U_xXtcLWLmqzWLleR_c54F/"
yemshanov_pineMap <- prepInputs(url = url_yemshanov,
                                targetFile = "Yemshanov_pine_map.flt",
                                alsoExtract = "similar",
                                fun = "raster::raster",
                                #rasterToMatch = rasterToMatch
                                #studyArea = sf::as_Spatial(studyArea),
                                destinationPath = inputsDir)

levelplot(yemshanov_pineMap, margin = list(median), maxpixels = 1e6, par.settings = myThemeGreen)
```

## Beaudoin *et al.* kNN pine maps

From @Beaudoin:2014cf.

### Download kNN

```{r kNN-download, eval = FALSE}
if (isTRUE(download.data)) {
  ## TODO: redo this using prepInputs()
  url <- "http://tree.nfis.org/"
  docFiles <- c("NFI_and_kNN_Mapping_20160628.docx",
                "FR_NFI_and_kNN_Mapping_20160628.docx",
                "cjfr-2013-0401suppl.pdf")
  dataFiles <- c("kNN-Species.tar",              ## percent pine layers
                 "kNN-StructureStandVolume.tar") ## age & total volume layers

  dl.data(paste0(url, c(docFiles, dataFiles)), inputsDir)

  ## percent pine layers
  untar(file.path(inputsDir, dataFiles[1]),
        files = c("NFI_MODIS250m_kNN_Species_Pinu_Ban_v0.zip",
                  "NFI_MODIS250m_kNN_Species_Pinu_Con_v0.zip"),
        exdir = path.expand(inputsDir))
  
  zipFile <- file.path(inputsDir, "NFI_MODIS250m_kNN_Species_Pinu_Ban_v0.zip")
  destDir2 <- file.path(tools::file_path_sans_ext(zipFile))
  unzip(zipFile, exdir = destDir2)
  
  zipFile <- file.path(inputsDir, "NFI_MODIS250m_kNN_Species_Pinu_Con_v0.zip")
  destDir2 <- file.path(tools::file_path_sans_ext(zipFile))
  unzip(zipFile, exdir = destDir2)
  
  ## age & total volume layers
  untar(file.path(inputsDir, dataFiles[2]),
        files = c("NFI_MODIS250m_kNN_Structure_Stand_Age_v0.zip",
                  "NFI_MODIS250m_kNN_Structure_Volume_Total_v0.zip"),
        exdir = path.expand(inputsDir))
  
  zipFile <- file.path(inputsDir, "NFI_MODIS250m_kNN_Structure_Stand_Age_v0.zip")
  destDir2 <- file.path(tools::file_path_sans_ext(zipFile))
  unzip(zipFile, exdir = destDir2)
  
  zipFile <- file.path(inputsDir, "NFI_MODIS250m_kNN_Structure_Volume_Total_v0.zip")
  destDir2 <- file.path(tools::file_path_sans_ext(zipFile))
  unzip(zipFile, exdir = destDir2)
}
```

### kNN percent pine (%)

```{r knn-raster-percentPine}
files <- file.path(inputsDir,
                   c("NFI_MODIS250m_kNN_Species_Pinu_Ban_v0",
                     "NFI_MODIS250m_kNN_Species_Pinu_Con_v0"),
                   c("NFI_MODIS250m_kNN_Species_Pinu_Ban_v0.tif",
                     "NFI_MODIS250m_kNN_Species_Pinu_Con_v0.tif"))
stopifnot(file.exists(files))

kNN_percentPine <- stack(x = files) %>% setNames(c("Jack_Pine", "Lodgepole_Pine"))

save(kNN_percentPine, file = file.path(inputsDir, "kNN_percentPine.Rdata"))
writeRaster(kNN_percentPine, filename = file.path(inputsDir, "kNN_percentPine.tif"), overwrite = TRUE)
file.copy(file.path(inputsDir, "kNN_percentPine.Rdata"),
          "~/Documents/GitHub/MPB/SpaDES/modules/mpbPine/data/kNN_pine_map.tif",
          overwrite = TRUE)
```

```{r knn-plot-percentPine}
levelplot(kNN_percentPine, margin = list(median), maxpixels = 1e6, par.settings = myThemeGreen)
```

### kNN pine volume (m^3/ha)

```{r knn-raster-volumePine}
f <- file.path(maps.dir, ifelse(._CFS_., "kNN/original", "kNN"),
               "kNN-StructureStandVolume",
               "NFI_MODIS250m_kNN_Structure_Volume_Total_v0",
               "NFI_MODIS250m_kNN_Structure_Volume_Total_v0.tif")
stopifnot(file.exists(f))

kNN_volumeTotal <- raster(f) %>% setNames("Volume_Total")
save(kNN_volumeTotal, file = file.path(inputsDir, "kNN_volumeTotal.Rdata"))

kNN_volumePine <- kNN_percentPine
kNN_volumePine[[1]] <- kNN_percentPine[[1]] * kNN_volumeTotal / 100 # Jack_Pine
kNN_volumePine[[2]] <- kNN_percentPine[[2]] * kNN_volumeTotal / 100 # Lodgepole_Pine
names(kNN_volumePine) <- names(kNN_percentPine)
save(kNN_volumePine, file = file.path(inputsDir, "kNN_volumePine.Rdata"))
writeRaster(kNN_volumePine, filename = file.path(inputsDir, "kNN_volumePine.tif"))
```

```{r knn-plot-volumePine}
## plot it
plot(kNN_volumePine_map)

## even better with rasterVis
#pdf(file.path("pine-maps", "pineVolume.pdf"), width = 12, height = 6)
levelplot(kNN_volumePine, margin = list(median), maxpixels = 1e6, par.settings = myThemeGreen)
#dev.off()
```

### Stand age (a)

```{r knn-raster-standAge}
kNN_standAge <- LandR::prepInputsStandAgeMap(
  startTime = 2010,
  ageUrl = paste0("http://ftp.maps.canada.ca/pub/nrcan_rncan/Forests_Foret/",
                                    "canada-forests-attributes_attributs-forests-canada/",
                                    "2001-attributes_attributs-2001/",
                                    "NFI_MODIS250m_2001_kNN_Structure_Stand_Age_v1.tif"),
  destinationPath = inputsDir,
  rasterToMatch = rasterToMatch,
  userTags = c("stable", currentModule(sim))
)
kNN_standAge[] <- LandR::asInteger(kNN_standAge[])
```

```{r knn-plot-standAge}
## plot it
plot(kNN_standAge)

## even better with rasterVis
levelplot(kNN_standAge, margin = list(median), maxpixels = 1e6, par.settings = myThemeBlue)
```

## AVI-derived pine maps for AB and SK

These were used to produce the pine maps in the latest MPB risk assessment [@Bleiker:2019].

```{r}
url_AB <- "https://drive.google.com/file/d/15EzncjIR_dn5v6hruoVbsUQVF706nTEL/"
AB <- Cache(prepInputs,
            url = url_AB,
            targetFile = "AB_PineVolumes_Lambert.gdb", ## TODO: not extracting correctly from zip. needs to be done manually :(
            alsoExtract = NA,
            archive = "AB_PineVolumes_Lambert.gdb.zip",
            fun = "sf::st_read", layer = "OVERSTOREY_PINE", ## TODO: UNDERSTOREY_PINE?
            overwrite = TRUE,
            destinationPath = inputsDir,
            cachePath = cacheDir) %>%
  fasterize::fasterize(., raster = rasterToMatch, field = "PCT_P")
AB <- AB * 10

url_SK <- "https://drive.google.com/file/d/1gpA9M4nhrnfIIvGQ7jcM9A7Fo-3MYpU1/"
SK <- Cache(prepInputs,
            url = url_SK,
            targetFile = "SK_INV_JPpct10_Lambert.tif",
            alsoExtract = "similar",
            fun = "raster::raster",
            rasterToMatch = rasterToMatch,
            destinationPath = inputsDir,
            cachePath = cacheDir)
SK <- SK * 10

AVI_percentPine <- raster::mosaic(AB, SK, fun = mean, na.rm = TRUE)
```

```{r avi-plot-percentPine}
## even better with rasterVis
levelplot(AVI_percentPine, margin = list(median), maxpixels = 1e6, par.settings = myThemeGreen)
```

## Other data sources?

```
Kyle Lochhead; Valerie LeMay; Gary Bull; Olaf Schwab; James Halperin, 2017-07-26, "Multivariate maps of forest attributes for management units in Canada's boreal forest", http://hdl.handle.net/11272/10515 V2 [Version]
```

# References

<!-- automatically generated using bibliography.bib -->
